Innovation and elaboration on the avian tree of life - supplementary materials

### 2022-11-09 Manuscript/main_flipped_narrative.tex
Update main_flipped_narrative.tex

\noindent {\normalsize \sc
Thomas Guillerme$^{1,*}$, 
Jen A. Bright$^{2}$,
Christopher R. Cooney${^1}$,
Emma C. Hughes${^1}$,
Zoe K. Varley${^3}$,
Natalie Cooper$^{3,+}$,
Andrew P. Beckerman$^{1,+}$,
and Gavin H. Thomas$^{1,4,+}$}\\
\noindent {\small \it 
$^1$School of Biosciences, University of Sheffield, Sheffield, S10 2TN, United Kingdom.\\
$^2$Department of Biological and Marine Sciences, University of Hull, Hull, HU6 7RX, United Kingdom.\\
$^3$Natural History Museum, Cromwell Road, London, SW7 5BD, United Kingdom.\\
$^4$Bird Group, Department of Life Sciences, the Natural History Museum at Tring, Tring, United Kingdom.\\}

@@

\noindent (Keywords: macroevolution, elaboration, innovation, variance-covariance matrices, line of least resistance, GLMM)\\

@@

Patterns of biological diversity across the tree of life are the result of millions of years of evolutionary history and are shaped by natural selection.
A long-standing proposal is that most morphological diversity at a macroevolutionary scale (among species) arises along a microevolutionary ``line of least resistance''.https://www.overleaf.com/project/62f128d3d7c36e4efd1e9b21
However, we frequently observe major shifts in phenotypes among lineages \cite{venditti2011multiple, pagel2022general, BurinWhales}, suggesting that species or clades may also evolve away from the line of least resistance \cite{renaud2006conserved}.
At the macroevolutionary scale, evolution along a common major axis has been termed elaboration whereas evolution away from the major axis is termed innovation \cite{endler2005animal}. 
The extent to which species and clades elaborate or innovate is poorly understood.
Of particular importance is the potential contribution of reorientations of the phenotypic trait space (innovation) among clades to contribute to the origins and maintenance of biodiversity across scales from the macro- to megaevolutionary scale.
Here we apply new multi-trait methods that model the reorientations of the trait space via the nested covariance structure of traits to evaluate the magnitude and distribution of elaboration and innovation in the evolution of bird beaks.
Our analyses show that elaboration dominates species-level phenotypic divergence, consistent with micro-evolutionary theory.
However, we also find that innovation is relatively common at the clade-level suggesting multiple routes to phenotypic innovation throughout avian evolutionary history.
These results show that although phenotypic elaboration and innovation occur across evolutionary scales, elaboration seems to dominate at macroevolutionary scales (i.e. species diverging within clades) and innovation is more prevalent at the megaevolutionary scale (nested patterns of clades diverging within clades).
This suggests that both innovation and elaboration are ubiquitous routes for phenotypic change at any scale and that macroevolutionary axes of multivariate evolution are frequently reoriented throughout the history of life, opening up new avenues for evolution to explore.

@@

% paragraph 1:  introducing the major axis of variation + novelty and innovation
Evolutionary theory predicts that over short, microevolutionary, timescales phenotypic evolution will tend to follow evolutionary lines of genetic least resistance \cite{schluter1996adaptive}.
Over longer timescales, the evolutionary line of least resistance is analogous to the major axis of phenotypic variation \cite{marroig2005size,fasanelli2022allometry}.
Numerous studies show that phenotypic divergence is biased along the genetic line of least resistance, and therefore reflects the major axis of phenotypic variation, over 1-2 million year timescales.
%TG: cite some of these numerous studies.
Recent evidence suggests that this stability can even extend over macroevolutionary timescales spanning tens of millions of years \cite{mcglothlin2018adaptive}.
However, while the line of least genetic resistance explicitly reflects genetic constraints, the macroevolutionary major axis of phenotypic variation is an emergent property of genomic and developmental constraints interacting with selection on a moving adaptive landscape \cite{jones2004evolution}.
The widespread evidence of major shifts in phenotypes across the tree of life \cite{pagel2022general,cooney2017mega,venditti2011multiple,khabbazian2016fast,smaers2021evolution} suggests a tension in the direction of phenotypic evolution across scales where macroevolutionary divergence away from the major axis of phenotypic variation results in megaevolutionary divergence leading to distinct phenotypes and/or directions of evolution.
Resolving this tension requires an understanding of the balance between deep-time reorientation of phenotypic major axes of evolution among clades and the directions of evolution of species within clades.

@@

\subsection{Elaboration dominates species-level divergence}
% paragraph 4: results for elaboration and innovation for species
We tested the expectation, derived from adaptive radiation theory, that species divergence is biased along lines of least resistance by calculating species-specific measures of innovation and elaboration.
If microevolutionary lines of least resistance are stable then we expect that species fall on a conserved major axis (elaboration) rather than fall of the line (innovation).
To do this we calculated the multilinear algebraic projection of each species phenotype onto three major axes of phenotypic variation (global phylogenetic, super-order, and order).
We found higher values of elaboration than innovation overall across all levels (Fig \ref{Fig:phylogeny}) and globaly more elaboration and innovation at the global phylogeny and the super-orders levels compared to the orders level.
Phylogenetic distributions of innovation and elaboration broadly hold whether comparisons are made at the order, super-order, or global phylogenetic level (Fig. \ref{Fig:phylogeny})%{#fig_correlations}),%TODO: maybe add to supp?
 and in further nested analysis of the order Passeriformes (Fig. S\ref{fig_ellipses_passeriformes} and S\ref{fig_phylogeny_passeriformes}).
However, the extent of both elaboration and innovation tends to reduce as we move from global phylogenetic to within-order comparison.
Broadly, and not surprisingly, this trend reflects the fact that species are more similar to the major axis of phenotypic variation of their close relatives than to the major axis of phenotypic variation of all birds.
The global phylogenetic major axis of phenotypic variation aligns closely with the major axis in the trait space and primarily describes variation between short, deep, and wide beaks at one extreme and long, shallow, and narrow beaks at the other.
Our results point to the possibility that within clades the major axes of phenotypic variation follow different directions.

Overall we found that, as expected from microevolutionary theory, elaboration is the dominant mode of divergence at the species level \ref{Fig:phylogeny}.
We find no consistent evidence that innovation is either positively correlated with, or trades-off with, elaboration. %(Fig. S{#fig_correlations}). %TODO: maybe add to supp? 
However, we observe various combinations of elaboration and innovation across birds.
Examples of high innovation and/or elaboration are often clustered within clades.
For example, compared to the global phylogenetic major axis of phenotypic variation, the Trochilidae (hummingbirds) in the order Apodiformes, have long narrow beaks and display high levels of elaboration and low innovation.
In contrast, the Bucerotiformes (hornbills), with large and deep beaks display high levels of innovation and low elaboration.
Other clades such as the Apodidae (swifts), in the order Apodiformes, and the distantly-related but convergent Hirundidae (swallows and martins) in the order Passeriformes, have exceptionally wide beaks relative to length, and display high levels of both elaboration and innovation.
The dominance of elaboration when measured at the species-level is consistent with idea of evolution along lines of least evolutionary resistance.
The frequent phylogenetic clustering of species that elaborate, innovate, or both, highlights the likelihood that major axes of phenotypic variation are not stable among clades.
This leads us to ask: if most organisms evolve incrementally along the major axis of phenotypic variation then how can we explain the staggering diversity of species morphologies across the tree of life?

@@

To investigate the tension implied by our observations of species-level elaboration and previously reported deep time jumps in phenotype, we measured the elaboration and innovation of different clades.

@@

\noindent \textbf{Figure \ref{Fig:ellipses}:}
Panel A) shows ellipses representing the scaled average posterior variance-covariance response from the pGLMM models for each super-order (coloured ellipses) compared to the global (all birds) phylogenetic component of the models (grey ellipses).
We scaled the ellipses so the length of the major axis of phenotypic variation of the clade ellipse is the same length as that of the global phylogenetic major axis ellipse (in eight dimensions).
The first inset ellipse plot shows the positions of all super-order ellipses relative to the global phylogenetic ellipse.
Subsequent inset plots show the results for each super-order. 
Inset barplots show the proportion of variance associated with each of the eight PC axes in shape morphospace.
The inset boxplots correspond to the elaboration (E) and innovation (I) scores for all 4000 posterior samples.
The dots represent the median elaboration and innovation values while the thick and dashed lines represent the 50\% and 95\% confidence intervals respectively.
These scores were calculated on the unscaled ellipses resulting in different scales of elaboration and innovation for each plot. %TG: TODO: add figure with unscalled ellipses in the supplementaries
Panel B) shows the results for each order.
A companion plot showing further nested structure within the Passeriformes is shown in Fig. S\ref{fig_ellipses_passeriformes}.

@@

\textbf{Reproducibility and repeatability.}
The raw and processed data used in this analysis is available from \href{https://figshare.com/articles/dataset/Innovation_and_elaboration_on_the_avian_tree_of_life/20480355}{figshare} \cite{fighsaredata}.
The code used to reproduce the figures and tables is available from \href{https://github.com/TGuillerme/elaboration_exploration_bird_beaks}{GitHub} \cite{githubrepo}.

>>>>

\noindent {\normalsize \sc
Thomas Guillerme$^{1,*}$, 
Jen A. Bright$^{2}$,
Christopher R. Cooney${^1}$,
Emma C. Hughes${^1}$,
Zoe K. Varley${^3}$,
Natalie Cooper$^{3,+}$,
Andrew P. Beckerman$^{1,+}$,
and Gavin H. Thomas$^{1,4,+}$}\\
\noindent {\small \it 
$^1$School of Biosciences, University of Sheffield, Sheffield, S10 2TN, United Kingdom.\\
$^2$School of Natural Science, University of Hull, Hull, HU6 7RX, United Kingdom.\\
$^3$Natural History Museum, Cromwell Road, London, SW7 5BD, United Kingdom.\\
$^4$Bird Group, Department of Life Sciences, the Natural History Museum at Tring, Tring, United Kingdom.\\}

@@

\noindent (Keywords: macroevolution, elaboration, innovation, variance-covariance matrices, major axis of phenotypic variation, GLMM)\\

@@

Patterns of biological diversity across the tree of life are the result of millions of years of evolutionary history and are shaped by natural selection.
A long-standing proposal is that most morphological diversity at a macroevolutionary scale (among species) arises along a microevolutionary ``line of least resistance''.
However, we frequently observe major shifts in phenotypes among lineages \cite{venditti2011multiple, pagel2022general, BurinWhales}, suggesting that species or clades may also evolve away from the line of least resistance \cite{renaud2006conserved}.
At the macroevolutionary scale, evolution along a common major axis has been termed elaboration whereas evolution away from the major axis is termed innovation \cite{endler2005animal}. 
The extent to which species and clades elaborate or innovate is poorly understood.
Of particular importance is the potential contribution of reorientations of the phenotypic trait space (innovation) among clades to contribute to the origins and maintenance of biodiversity across scales from the macro- to megaevolutionary scale.
Here we apply new multi-trait methods that model the reorientations of the trait space via the nested covariance structure of traits to evaluate the magnitude and distribution of elaboration and innovation in the evolution of bird beaks.
Our analyses show that elaboration dominates species-level phenotypic divergence, consistent with micro-evolutionary theory.
However, we also find that innovation is relatively common at the clade-level suggesting multiple routes to phenotypic innovation throughout avian evolutionary history.
These results show that although phenotypic elaboration and innovation occur across evolutionary scales, elaboration seems to dominate at macroevolutionary scales (i.e. species diverging within clades) and innovation is more prevalent at the megaevolutionary scale (nested patterns of clades diverging within clades).
This suggests that both innovation and elaboration are ubiquitous routes for phenotypic change at any scale and that macroevolutionary axes of multivariate evolution are frequently reoriented throughout the history of life, opening up new avenues for evolution to explore.

@@

% paragraph 1:  introducing the major axis of variation + novelty and innovation
Evolutionary theory predicts that over short, microevolutionary, timescales phenotypic evolution will tend to follow evolutionary lines of genetic least resistance \cite{schluter1996adaptive}.
Over longer timescales, the evolutionary line of least resistance is analogous to the major axis of phenotypic variation \cite{marroig2005size,fasanelli2022allometry}.
Numerous studies show that phenotypic divergence is biased along the genetic line of least resistance, and therefore reflects the major axis of phenotypic variation, over 1-2 million year timescales.
%TG: cite some of these numerous studies.
Recent evidence suggests that this stability can even extend over macroevolutionary timescales spanning tens of millions of years \cite{mcglothlin2018adaptive}.
However, while the line of least genetic resistance explicitly reflects genetic constraints, the macroevolutionary major axis of phenotypic variation is an emergent property of genomic and developmental constraints interacting with selection on a moving adaptive landscape \cite{jones2004evolution}.
The widespread evidence of major shifts in phenotypes across the tree of life \cite{pagel2022general,cooney2017mega,venditti2011multiple,khabbazian2016fast,smaers2021evolution} suggests a confict in the direction of phenotypic evolution across scales where macroevolutionary divergence away from the major axis of phenotypic variation results in megaevolutionary divergence leading to distinct phenotypes and/or directions of evolution.
Resolving this confict requires an understanding of the balance between deep-time reorientation of phenotypic major axes of evolution among clades and the directions of evolution of species within clades.

@@

\subsection{Elaboration dominates species-level divergence}
% paragraph 4: results for elaboration and innovation for species
We tested the expectation, derived from adaptive radiation theory, that species divergence is biased along lines of least resistance by calculating species-specific measures of innovation and elaboration.
If microevolutionary lines of least resistance are stable then we expect that species fall on a conserved major axis (elaboration) rather than fall off the line (innovation).
To do this we calculated the multilinear algebraic projection of each species phenotype onto three major axes of phenotypic variation (global phylogenetic, super-order, and order).
We found higher values of elaboration than innovation overall across all levels (Fig \ref{Fig:phylogeny}) and globaly more elaboration and innovation at the global phylogeny and the super-orders levels compared to the orders level.
Phylogenetic distributions of innovation and elaboration broadly hold whether comparisons are made at the order, super-order, or global phylogenetic level (Fig. \ref{Fig:phylogeny})%{#fig_correlations}),%TODO: maybe add to supp?
 and in further nested analysis of the order Passeriformes (Fig. S\ref{fig_ellipses_passeriformes} and S\ref{fig_phylogeny_passeriformes}).
However, the extent of both elaboration and innovation tends to reduce as we move from global phylogenetic to within-order comparison.
Broadly, and not surprisingly, this trend reflects the fact that species are more similar to the major axis of phenotypic variation of their close relatives than to the major axis of phenotypic variation of all birds.
The global phylogenetic major axis of phenotypic variation aligns closely with the major axis in the trait space and primarily describes variation between short, deep, and wide beaks at one extreme and long, shallow, and narrow beaks at the other.
Our results point to the possibility that within clades the major axes of phenotypic variation follow different directions.

Overall we found that, as expected from microevolutionary theory, elaboration is the dominant mode of divergence at the species level (Fig. \ref{Fig:phylogeny}).
We find no consistent evidence that innovation is either positively correlated with, or trades-off with, elaboration. %(Fig. S{#fig_correlations}). %TODO: maybe add to supp? 
However, we observe various combinations of elaboration and innovation across birds.
Examples of high innovation and/or elaboration are often clustered within clades.
For example, compared to the global phylogenetic major axis of phenotypic variation, the Trochilidae (hummingbirds) in the order Apodiformes, have long narrow beaks and display high levels of elaboration and low innovation.
In contrast, the Bucerotiformes (hornbills), with large and deep beaks display high levels of innovation and low elaboration.
Other clades such as the Apodidae (swifts), in the order Apodiformes, and the distantly-related but convergent Hirundidae (swallows and martins) in the order Passeriformes, have exceptionally wide beaks relative to length, and display high levels of both elaboration and innovation.
The dominance of elaboration when measured at the species-level is consistent with idea of evolution along lines of least evolutionary resistance.
The frequent phylogenetic clustering of species that elaborate, innovate, or both, highlights the likelihood that major axes of phenotypic variation are not stable among clades.
This leads us to ask: if most organisms evolve incrementally along the major axis of phenotypic variation then how can we explain the staggering diversity of species morphologies across the tree of life?

@@

To investigate the conflict implied by our observations of species-level elaboration and previously reported deep time jumps in phenotype, we measured the elaboration and innovation of different clades.

@@

\noindent \textbf{Figure \ref{Fig:ellipses}:}
Panel A) shows ellipses representing the scaled average posterior variance-covariance response from the pGLMM models for each super-order (coloured ellipses) compared to the global (all birds) phylogenetic component of the models (grey ellipses).
Each ellipses represents a two dimensional projection of the average traits variance co-variance in the global bird beak shapespace.
The ellipses are centred on the centroid of each clade they represent.
Their direction represents the major direction of the traits covariance in two dimensions.
We scaled the ellipses so the length of the major axis of phenotypic variation of the clade ellipse is the same length as that of the global phylogenetic major axis ellipse (in eight dimensions).
The first inset ellipse plot shows the positions of all super-order ellipses relative to the global phylogenetic ellipse.
Subsequent inset plots show the results for each super-order. 
Inset barplots show the proportion of variance associated with each of the eight PC axes in shape morphospace.
The inset boxplots correspond to the elaboration (E) and innovation (I) scores for all 4000 posterior samples.
The dots represent the median elaboration and innovation values while the thick and dashed lines represent the 50\% and 95\% confidence intervals respectively.
These scores were calculated on the unscaled ellipses resulting in different scales of elaboration and innovation for each plot.
Panel B) shows the results for each order.
A companion plot showing further nested structure within the Passeriformes is shown in Fig. S\ref{fig_ellipses_passeriformes}.

@@

\textbf{Reproducibility and repeatability.}
The raw and processed data used in this analysis is available from \href{https://figshare.shef.ac.uk/account/articles/21526146}{figshare} \cite{fighsaredata}.
The code used to reproduce the figures and tables is available from \href{https://github.com/TGuillerme/elaboration_exploration_bird_beaks}{GitHub} \cite{githubrepo}.
